---
title: "Non-linear models"
author: "Brooke Anderson"
date: "March 10, 2016"
output: beamer_presentation
---

```{r warning = FALSE, message = FALSE, echo = FALSE}
#knitr::opts_knit$set(root.dir = "..") # Reset root directory for analysis
library(lubridate) # To help handle dates
library(dplyr) # Data wrangling
library(ggplot2) # Plotting
library(gridExtra) # To arrange multiple ggplot objects in one graph
library(splines) # To fit splines
library(gam)
library(caret)

train <- read.csv("../data/train.csv", as.is = TRUE) # `as.is` so `datetime` comes in as
                                                  # character, not factor
test <- read.csv("../data/test.csv", as.is = TRUE)

train <- mutate(train,
                datetime = ymd_hms(datetime),
                year = year(datetime),
                hour = hour(datetime),
                month = month(datetime),
                yday = yday(datetime),
                wday = factor(as.character(wday(datetime, label = TRUE))),
                season = factor(season, levels = c(1, 2, 3, 4),
                                labels = c("Spring", "Summer", "Fall", "Winter")),
                workingday = factor(workingday, levels = c(0, 1),
                                    labels = c("Holiday / weekend",
                                               "Working day")))
test  <- mutate(test,
                datetime = ymd_hms(datetime),
                year = year(datetime),
                hour = hour(datetime),
                month = month(datetime),
                yday = yday(datetime),
                wday = factor(as.character(wday(datetime, label = TRUE))),
                season = factor(season, levels = c(1, 2, 3, 4),
                                labels = c("Spring", "Summer", "Fall", "Winter")),
                workingday = factor(workingday, levels = c(0, 1),
                                    labels = c("Holiday / weekend",
                                               "Working day")))

rmsle <- function(train_preds, actual_preds){
  log_p_1 <- log(train_preds + 1)
  log_a_1 <- log(actual_preds + 1)
  sle <- (log_p_1 - log_a_1)^2
  rmsle <- sqrt(mean(sle))
  return(rmsle)
}

write_test_preds <- function(test_preds, mod_name){
  out_file <- data.frame(datetime = as.character(test$datetime),
                         count = test_preds)
  out_name <- paste0("../test_predictions/", mod_name, ".csv")
  write.csv(out_file, file = out_name, row.names = FALSE)
}
```

## Spline of hour

```{r, fig.show = "hide"}
ex_1 <- glm(count ~ ns(hour, 6), data = train,
            family = quasipoisson)
newdata <- data.frame(hour = 0:23)
newdata$y_hat <- predict(ex_1, newdata = newdata,
                         type = "response")

ggplot(newdata, aes(x = hour, y = y_hat)) + 
  geom_point(data = train, aes(x = hour, y = count),
             alpha = 0.1) + 
  geom_line(color = "red", size = 2) + 
  theme_minimal()
```

## Spline of hour

Degrees of freedom: 6

```{r, echo = FALSE}
ex_1 <- glm(count ~ ns(hour, 6), data = train,
            family = quasipoisson)
newdata <- data.frame(hour = 0:23)
newdata$y_hat <- predict(ex_1, newdata = newdata,
                         type = "response")

ggplot(newdata, aes(x = hour, y = y_hat)) + 
  geom_point(data = train, aes(x = hour, y = count),
             alpha = 0.1) + 
  geom_line(color = "red", size = 2) + 
  theme_minimal()
```

## Spline of hour

Degrees of freedom: 3

```{r, echo = FALSE}
ex_1 <- glm(count ~ ns(hour, 3), data = train,
            family = quasipoisson)
newdata <- data.frame(hour = 0:23)
newdata$y_hat <- predict(ex_1, newdata = newdata,
                         type = "response")

ggplot(newdata, aes(x = hour, y = y_hat)) + 
  geom_point(data = train, aes(x = hour, y = count),
             alpha = 0.1) + 
  geom_line(color = "red", size = 2) + 
  theme_minimal()
```

## Spline of hour

Degrees of freedom: 15

```{r, echo = FALSE}
ex_1 <- glm(count ~ ns(hour, 15), data = train,
            family = quasipoisson)
newdata <- data.frame(hour = 0:23)
newdata$y_hat <- predict(ex_1, newdata = newdata,
                         type = "response")

ggplot(newdata, aes(x = hour, y = y_hat)) + 
  geom_point(data = train, aes(x = hour, y = count),
             alpha = 0.1) + 
  geom_line(color = "red", size = 2) + 
  theme_minimal()
```

## Spline of day of year

```{r, fig.show = "hide"}
ex_1 <- glm(count ~ ns(yday, 6), data = train,
            family = quasipoisson)
newdata <- data.frame(yday = 1:366)
newdata$y_hat <- predict(ex_1, newdata = newdata,
                         type = "response")

ggplot(newdata, aes(x = yday, y = y_hat)) + 
  geom_point(data = train, aes(x = yday, y = count),
             alpha = 0.1) + 
  geom_line(color = "red", size = 2) + 
  theme_minimal()
```

## Spline of day of year

Degrees of freedom: 6

```{r, echo = FALSE}
ex_1 <- glm(count ~ ns(yday, 6), data = train,
            family = quasipoisson)
newdata <- data.frame(yday = 1:366)
newdata$y_hat <- predict(ex_1, newdata = newdata,
                         type = "response")

ggplot(newdata, aes(x = yday, y = y_hat)) + 
  geom_point(data = train, aes(x = yday, y = count),
             alpha = 0.1) + 
  geom_line(color = "red", size = 2) + 
  theme_minimal()
```

## Spline of day of year

Degrees of freedom: 12

```{r, echo = FALSE}
ex_1 <- glm(count ~ ns(yday, 12), data = train,
            family = quasipoisson)
newdata <- data.frame(yday = 1:366)
newdata$y_hat <- predict(ex_1, newdata = newdata,
                         type = "response")

ggplot(newdata, aes(x = yday, y = y_hat)) + 
  geom_point(data = train, aes(x = yday, y = count),
             alpha = 0.1) + 
  geom_line(color = "red", size = 2) + 
  theme_minimal()
```

## Model 

On Kaggle: 0.77821.

```{r}
mod_1 <- glm(count ~ ns(hour, 12) + ns(yday, 6),
             data = train, family = quasipoisson)

train_preds <- predict(mod_1, type = "response")
actual_preds <- train$count 
rmsle(train_preds, actual_preds)

test_preds <- predict(mod_1, newdata = test,
                      type = "response")
write_test_preds(test_preds, mod_name = "gam_1")
```

## Spline of hour

```{r, fig.show = "hide"}
ex_2 <- glm(count ~ workingday + ns(hour, 12),
            data = train,
            family = quasipoisson)
newdata <- data.frame(hour = rep(0:23, 2),
            workingday = rep(unique(train$workingday),
                                       each = 24))
newdata$y_hat <- predict(ex_2, newdata = newdata,
                         type = "response")

ggplot(newdata, aes(x = hour, y = y_hat)) + 
  geom_point(data = train, aes(x = hour, y = count),
             alpha = 0.1) + 
  geom_line(color = "red", size = 2) + 
  facet_wrap(~ workingday, ncol = 1) + 
  theme_minimal()
```

## Spline of hour

```{r, echo = FALSE}
ex_2 <- glm(count ~ workingday + ns(hour, 12), data = train,
            family = quasipoisson)
newdata <- data.frame(hour = rep(0:23, 2),
                      workingday = rep(unique(train$workingday),
                                       each = 24))
newdata$y_hat <- predict(ex_2, newdata = newdata,
                         type = "response")

ggplot(newdata, aes(x = hour, y = y_hat)) + 
  geom_point(data = train, aes(x = hour, y = count),
             alpha = 0.1) + 
  geom_line(color = "red", size = 2) + 
  facet_wrap(~ workingday, ncol = 1) + 
  theme_minimal()
```

## Spline of hour

```{r, fig.show = "hide"}
ex_2 <- glm(count ~ workingday*ns(hour, 12),
            data = train,
            family = quasipoisson)
newdata <- data.frame(hour = rep(0:23, 2),
            workingday = rep(unique(train$workingday),
                                       each = 24))
newdata$y_hat <- predict(ex_2, newdata = newdata,
                         type = "response")

ggplot(newdata, aes(x = hour, y = y_hat)) + 
  geom_point(data = train, aes(x = hour, y = count),
             alpha = 0.1) + 
  geom_line(color = "red", size = 2) + 
  facet_wrap(~ workingday, ncol = 1) + 
  theme_minimal()
```

## Spline of hour

```{r, echo = FALSE}
ex_2 <- glm(count ~ workingday*ns(hour, 12), data = train,
            family = quasipoisson)
newdata <- data.frame(hour = rep(0:23, 2),
                      workingday = rep(unique(train$workingday),
                                       each = 24))
newdata$y_hat <- predict(ex_2, newdata = newdata,
                         type = "response")

ggplot(newdata, aes(x = hour, y = y_hat)) + 
  geom_point(data = train, aes(x = hour, y = count),
             alpha = 0.1) + 
  geom_line(color = "red", size = 2) + 
  facet_wrap(~ workingday, ncol = 1) + 
  theme_minimal()
```

## Model 

On Kaggle: 0.60629.

```{r}
mod_2 <- glm(count ~ workingday * ns(hour, 12) + 
               ns(yday, 6),
             data = train, family = quasipoisson)

train_preds <- predict(mod_2, type = "response")
actual_preds <- train$count 
rmsle(train_preds, actual_preds)

test_preds <- predict(mod_2, newdata = test,
                      type = "response")
write_test_preds(test_preds, mod_name = "gam_2")
```

## Spline of hour

```{r}
ex_3 <- glm(count ~ year + workingday*ns(hour, 12),
            data = train,
            family = quasipoisson)

new_hour <- 0:23
new_workingday <- levels(train$workingday)
new_year <- unique(train$year)

newdata <- expand.grid(new_hour, new_workingday, new_year)
colnames(newdata) <- c("hour", "workingday", "year")
head(newdata, 3)
```

## Spline of hour

```{r fig.show = "hide"}
newdata$y_hat <- predict(ex_3, newdata = newdata,
                         type = "response")

ggplot(newdata, aes(x = hour, y = y_hat)) + 
  geom_point(data = train, aes(x = hour, y = count),
             alpha = 0.1) + 
  geom_line(color = "red", size = 2) + 
  facet_grid(year~ workingday) + 
  theme_minimal()
```

## Spline of hour

```{r echo = FALSE}
newdata$y_hat <- predict(ex_3, newdata = newdata,
                         type = "response")

ggplot(newdata, aes(x = hour, y = y_hat)) + 
  geom_point(data = train, aes(x = hour, y = count),
             alpha = 0.1) + 
  geom_line(color = "red", size = 2) + 
  facet_grid(year~ workingday) + 
  theme_minimal()
```

## Spline of hour

```{r}
ex_4 <- glm(count ~ year*workingday*ns(hour, 12),
            data = train,
            family = quasipoisson)

new_hour <- 0:23
new_workingday <- levels(train$workingday)
new_year <- unique(train$year)

newdata <- expand.grid(new_hour, new_workingday, new_year)
colnames(newdata) <- c("hour", "workingday", "year")

newdata$y_hat <- predict(ex_4, newdata = newdata,
                         type = "response")
```

## Spline of hour

```{r echo = FALSE}
ggplot(newdata, aes(x = hour, y = y_hat)) + 
  geom_point(data = train, aes(x = hour, y = count),
             alpha = 0.1) + 
  geom_line(color = "red", size = 2) + 
  facet_grid(year~ workingday) + 
  theme_minimal()
```

## Model 

On Kaggle: 0.54972.

```{r}
mod_3 <- glm(count ~ workingday * ns(hour, 12) + 
               ns(yday, 6) + year,
             data = train, family = quasipoisson)

train_preds <- predict(mod_3, type = "response")
actual_preds <- train$count 
rmsle(train_preds, actual_preds)

test_preds <- predict(mod_3, newdata = test,
                      type = "response")
write_test_preds(test_preds, mod_name = "gam_3")
```

## Spline of temperature

```{r}
ex_5 <- glm(count ~ ns(temp, 5),
            data = train,
            family = quasipoisson)

newdata <- data.frame(temp = seq(from = 
                                floor(min(train$temp)),
                                 to = max(train$temp)))

newdata$y_hat <- predict(ex_5, newdata = newdata,
                         type = "response")
```

## Spline of temperature

```{r echo = FALSE}
ggplot(newdata, aes(x = temp, y = y_hat)) + 
  geom_point(data = train, aes(x = temp, y = count),
             alpha = 0.1) + 
  geom_line(color = "red", size = 2) + 
  theme_minimal()
```

## Spline of temperature

Degrees of freedom: 10.

```{r echo = FALSE}
ex_5 <- glm(count ~ ns(temp, 10),
            data = train,
            family = quasipoisson)

newdata <- data.frame(temp = floor(min(train$temp)):max(train$temp))

newdata$y_hat <- predict(ex_5, newdata = newdata,
                         type = "response")

ggplot(newdata, aes(x = temp, y = y_hat)) + 
  geom_point(data = train, aes(x = temp, y = count),
             alpha = 0.1) + 
  geom_line(color = "red", size = 2) + 
  theme_minimal()
```

## Model 

On Kaggle: 0.51936.

```{r}
mod_4 <- glm(count ~ workingday * ns(hour, 12) + 
               ns(yday, 6) + year + 
               ns(temp, 10),
             data = train, family = quasipoisson)

train_preds <- predict(mod_4, type = "response")
actual_preds <- train$count 
rmsle(train_preds, actual_preds)

test_preds <- predict(mod_4, newdata = test,
                      type = "response")
write_test_preds(test_preds, mod_name = "gam_4")
```

## Model 

On Kaggle: 0.46740.

```{r}
mod_5 <- glm(count ~ season * workingday * ns(hour, 12) + 
               ns(yday, 6) + year + 
               ns(temp, 10) + weather + wday +
               ns(humidity, 5),
             data = train, family = quasipoisson)

train_preds <- predict(mod_5, type = "response")
actual_preds <- train$count 
rmsle(train_preds, actual_preds)

test_preds <- predict(mod_5, newdata = test,
                      type = "response")
write_test_preds(test_preds, mod_name = "gam_5")
```

## Spline of hour

```{r}
ex_5a <- glm(registered ~ year*workingday*ns(hour, 12),
            data = train,
            family = quasipoisson)
ex_5b <- glm(casual ~ year*workingday*ns(hour, 12),
            data = train,
            family = quasipoisson)

new_hour <- 0:23
new_workingday <- levels(train$workingday)
new_year <- unique(train$year)
```

## Spline of hour

```{r}
newdata <- expand.grid(new_hour, new_workingday,
                       new_year)
colnames(newdata) <- c("hour", "workingday", "year")

newdata$y_hat_registered <- predict(ex_5a,
                                    newdata = newdata,
                                    type = "response")
newdata$y_hat_casual <- predict(ex_5b,
                                newdata = newdata,
                                type = "response")
```

## Spline of hour

Registered: 

```{r echo = FALSE}
ggplot(newdata, aes(x = hour, y = y_hat_registered)) + 
  geom_point(data = train, aes(x = hour, y = registered),
             alpha = 0.1) + 
  geom_line(color = "red", size = 2) + 
  facet_grid(year~ workingday) + 
  theme_minimal()
```

## Spline of hour

Casual: 

```{r echo = FALSE}
ggplot(newdata, aes(x = hour, y = y_hat_casual)) + 
  geom_point(data = train, aes(x = hour, y = casual),
             alpha = 0.1) + 
  geom_line(color = "red", size = 2) + 
  facet_grid(year~ workingday) + 
  theme_minimal()
```

## Model 

```{r}
mod_6a <- glm(registered ~ season * workingday * 
                ns(hour, 12) + 
               ns(yday, 6) + year + 
               ns(temp, 10) + weather + wday +
               ns(humidity, 5),
             data = train, family = quasipoisson)
mod_6b <- glm(casual ~ season * workingday * 
                ns(hour, 12) + 
               ns(yday, 6) + year + 
               ns(temp, 10) + weather + wday +
               ns(humidity, 5),
             data = train, family = quasipoisson)
```

## Model 

On Kaggle: 0.47690.

```{r}
train_predsa <- predict(mod_6a, type = "response")
train_predsb <- predict(mod_6b, type = "response")
train_preds <- train_predsa + train_predsb
actual_preds <- train$count 
rmsle(train_preds, actual_preds)

test_predsa <- predict(mod_6a, newdata = test,
                      type = "response")
test_predsb <- predict(mod_6b, newdata = test,
                      type = "response")
test_preds <- test_predsa + test_predsb
write_test_preds(test_preds, mod_name = "gam_6")
```

## K-nearest neighbors

```{r echo = FALSE}
rmsle_fun <- function(data, lev = NULL,
                      model = NULL, ...){
  log_p_1 <- log(data$pred + 1)
  log_a_1 <- log(data$obs + 1)
  sle <- (log_p_1 - log_a_1)^2
  rmsle <- sqrt(mean(sle))
  names(rmsle) <- "rmsle"
  return(rmsle)
}

fitControl <- trainControl(method = "cv",
                           number = 5,
                           summaryFunction = rmsle_fun)
```

```{r}
set.seed(825)
mod_1 <- train(count ~ temp + hour + 
                 workingday + year,
               data = train,
               method = "knn",
               trControl = fitControl,
               metric = "rmsle",
               maximize = FALSE,
               preProcess = c("center", "scale",
                              "spatialSign"),
               tuneLength = 5)

train_preds <- predict(mod_1,
                       newdata = train)
rmsle(train_preds, train$count)
```

## K-nearest neighbors

```{r}
plot(mod_1)
```

## K-nearest neighbors

On Kaggle: 0.49535.

```{r}
test_preds <- predict(mod_1, newdata = test)
write_test_preds(test_preds, mod_name = "knn")
```

## K-nearest neighbors

```{r}
set.seed(825)
mod_2 <- train(count ~ temp + factor(hour) + workingday + year,
               data = train,
               method = "knn",
               trControl = fitControl,
               metric = "rmsle",
               maximize = FALSE,
               preProcess = c("center", "scale",
                              "spatialSign"),
               tuneLength = 5)

train_preds <- predict(mod_2, newdata = train)
rmsle(train_preds, train$count)
```

## K-nearest neighbors

```{r}
plot(mod_2)
```

## K-nearest neighbors

On Kaggle: 0.48618.

```{r}
test_preds <- predict(mod_2, newdata = test)
write_test_preds(test_preds, mod_name = "knn2")
```

## SVM

```{r echo = FALSE}
rmsle_fun <- function(data, lev = NULL,
                      model = NULL, ...){
  log_p_1 <- log(exp(data$pred) + 1)
  log_a_1 <- log(data$obs + 1)
  sle <- (log_p_1 - log_a_1)^2
  rmsle <- sqrt(mean(sle))
  names(rmsle) <- "rmsle"
  return(rmsle)
}

fitControl <- trainControl(method = "cv",
                           number = 5,
                           summaryFunction = rmsle_fun)
```

```{r message = FALSE, warning = FALSE}
set.seed(825)
mod_3 <- train(log(count) ~ temp + hour + 
                 workingday + year,
               data = train,
               method = "svmRadial",
               trControl = fitControl,
               metric = "rmsle",
               maximize = FALSE,
               preProcess = c("center", "scale",
                              "spatialSign"),
               tuneLength = 5)

train_preds <- exp(predict(mod_3,
                           newdata = train))
rmsle(train_preds, train$count)
```

## SVM

On Kaggle: 0.77061.

```{r}
test_preds <- exp(predict(mod_3, newdata = test))
write_test_preds(test_preds, mod_name = "svm")
```


```{r message = FALSE, warning = FALSE}
set.seed(825)
mod_4 <- train(log(count) ~ ns(temp, 10) + 
                 ns(hour, 10) * workingday + year,
               data = train,
               method = "svmRadial",
               trControl = fitControl,
               metric = "rmsle",
               maximize = FALSE,
               preProcess = c("center", "scale",
                              "spatialSign"),
               tuneLength = 5)

train_preds <- exp(predict(mod_4,
                           newdata = train))
rmsle(train_preds, train$count)
```

