---
title: "`data.table`: Fast manipulation of large datasets in R"
author: "Brooke Anderson"
date: "April 25, 2016"
fontsize: 10pt
output: beamer_presentation
---

## `data.table`

`data.table` is a package in R that can efficiently read in and manipulate large datasets. It offers a **substantial** speed improvement over the classic `data.frame` when working with large datasets. 

## Example: US precipitation

As an example, I have a file with daily precipitation measures for every US county from 1979 through 2011: 

- 365 days * 33
- ~3,000 counties

This file has > 37,000,000 lines. The total file size is 2.26 GB. 

## Reading in a large text file

`fread` is the `data.table` equivalent of the `read.table` family of functions: 

```{r echo = FALSE}
precip_dir <- "/Users/brookeanderson/Documents/CSU2016/hurricaneproject/hurricaneexposure/data-raw/"
```

```{r message = FALSE}
library(data.table)
system.time(precip <- fread(paste0(precip_dir,
                       "nasa_precip_export_2.txt"),
                      header = TRUE,
                      select = c("county",
                                 "year_month_day",
                                 "precip"),
                      verbose = FALSE))
dim(precip)
```

## Reading in a large text file

`fread` can also read a file directly from http and https URLs, if you'd prefer to not save the flat file locally. 

## Manipulating a `data.table`

The `data.table` class has a series of conventions for summarizing and indexing that runs much, much faster than if you tried to use "classic" R functions. 

The general form is: 

```{r eval = FALSE}
precip[i, j, by]
```

where `i` filters by row, `j` selects or calculates on columns, and `by` groups by some grouping variable when selecting or calculating using columns.

## Manipulating a `data.table`

You can use the first element to filter to certain rows. For example, to pull out just values for Larimer County, CO, run: 

```{r}
precip[county == 8069 & 
         year_month_day %in% 
         c(19970727, 19970728), ]
```

## Manipulating a `data.table`

You can use the `order` function in the first element to sort the data: 

```{r}
head(precip[order(-precip), ])
```

## Manipulating a `data.table`

You can run calculations on columns using the second element: 

```{r}
precip[ , max(precip)]
precip[ , quantile(precip,
                   probs = c(0.99, 0.999,
                             0.9999))]
```

## Manipulating a `data.table`

You can combine filtering by rows and calculating on columns. For example, to figure out how many counties there were in 2011: 

```{r}
precip[year_month_day == 20110101,
       length(precip)]
```

*Note*: If you want to count rows, you can also use `.N`:

```{r}
precip[year_month_day == 20110101,
       .N]
```

## Grouped analysis

You can also group by a variable before you run an analysis. For example, to get the highest recorded precipitation in each county: 

```{r}
highest_precip <- precip[ , .(max.precip = max(precip)),
                          by = .(county)]
head(highest_precip, 3)
```

## Highest precipitation by county

```{r echo = FALSE, warning = FALSE, message = FALSE, fig.width = 8}
library(dplyr)
highest_precip <- rename(highest_precip, 
                          region = county, 
                          value = max.precip)

library(choroplethr)
county_choropleth(highest_precip)
```

## Chaining operation with `data.table`

If you want to, you can chain together several operations. For example, to determine the number of days over the 99.9th percentile in each county:

```{r}
extreme_precip <- precip[ , .N, .(precip > 
                                    quantile(precip,
                                             probs = 0.999),
                                 county)][
                          precip == TRUE, 
                        ]
```

## Extreme precipitation by county

```{r echo = FALSE, warning = FALSE, message = FALSE, fig.width = 8}
extreme_precip <- rename(extreme_precip, 
                          region = county, 
                          value = N)
county_choropleth(extreme_precip)
```

## Chaining operation with `data.table`

To plot trends by month within states: 

```{r}
ts_precip <- precip[ , .(precip = precip,
                         state = substring(sprintf("%05d",
                                                   county),
                                           1, 2),
                         month = as.numeric(
                           substring(year_month_day,
                                           5, 6)))][
                     , .(precip = mean(precip)),
                     keyby = .(state, month)                        
                     ]
```

## Precipitation by month and state

```{r echo = FALSE, fig.width = 8}
library(dplyr)
ts_precip2 <- filter(as.data.frame(ts_precip), 
                     state %in% c("22", "08", "51", "32")) %>%
  mutate(state = factor(state,
                        levels = c("22", "08", "51", "32"),
                        labels = c("Louisiana", "Colorado",
                                   "Virginia", "Nevada")))
library(ggplot2)
ggplot(ts_precip2, aes(x = month, y = precip)) + 
  geom_line() + 
  facet_wrap(~ state, ncol = 2)
```




